---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, include = FALSE}
knitr::opts_chunk$set(
    collapse = TRUE,
    comment = "#>",
    fig.path = "man/figures/README-",
    out.width = "100%"
)
```

# VectraPolarisData

<!-- badges: start -->
<!-- badges: end -->

The `VectraPolarisData` package provides a collection of [Vectra 3 and Vectra Polaris](https://www.akoyabio.com/phenoptics/mantra-vectra-instruments/) multiplex imaging datasets. Data cover various pathologies and tissues, and are provided as objects of class `SpatialExperiment`. 


## Installation instructions

The package is currently only available through GitHub, and should be installed using:

```{r 'install', eval = FALSE}
if (!requireNamespace("BiocManager", quietly = TRUE)) {
    install.packages("BiocManager")
}


BiocManager::install("julia-wrobel/VectraPolarisData")
#BiocManager::install("VectraPolarisData")
```

## Example

Here you can load the daa.

```{r example, eval = requireNamespace('VectraPolarisData')}
library("VectraPolarisData")
## basic example code
```


## Citation

Below is the citation output from using `citation('VectraPolarisData')` in R. Please
run this yourself to check for any updates on how to cite __VectraPolarisData__.

```{r 'citation', eval = requireNamespace('VectraPolarisData')}
print(citation('VectraPolarisData'), bibtex = TRUE)
```

Please note that the `VectraPolarisData` was only made possible thanks to many other R and bioinformatics software authors, which are cited either in the vignettes and/or the paper(s) describing this package.

