---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, include = FALSE}
knitr::opts_chunk$set(
    collapse = TRUE,
    comment = "#>",
    fig.path = "man/figures/README-",
    out.width = "100%"
)
```

# VectraPolarisData

<!-- badges: start -->
<!-- badges: end -->

The `VectraPolarisData` package provides two large multiplex immunofluorescence datasets, collected using [Vectra 3 and Vectra Polaris](https://www.akoyabio.com/phenoptics/mantra-vectra-instruments/) instruments. Data included a study of lung cancer and a study of ovarian cancer, and are provided as objects of class `SpatialExperiment`. 


## Installation instructions

Now available in Bioconductor version 3.15! To install, use:

```{r 'install', eval = FALSE}
if (!require("BiocManager", quietly = TRUE))
    install.packages("BiocManager")
}

BiocManager::install("VectraPolarisData")
```

## Example

Here you can load data from the package. Data from the [lung cancer study](https://www.jto.org/article/S1556-0864(21)02175-4/fulltext) are loaded below.

```{r example1, eval = requireNamespace('VectraPolarisData')}
library(VectraPolarisData)
spe_lung <- HumanLungCancerV3()
spe_lung
```


Data from the [ovarian cancer study](https://aacrjournals.org/mcr/article/19/12/1973/675069/The-Spatial-Context-of-Tumor-Infiltrating-Immune) are loaded below.

```{r example2, eval = requireNamespace('VectraPolarisData')}
spe_ovarian <- HumanOvarianCancerVP()
spe_ovarian
```



These datasets are provided as `SpatialExperiment` class objects. More detailed information about these datasets is available in the package vignette.
