---
title: "TENxVisiumData"
author:
- name: Julia Wrobel
  affiliation: Department of Biostatistics and Informatics, Colorado School of Public Health
output:
  BiocStyle::html_document:
    toc_float: true
package: VectraPolarisData
abstract: |
  The VectraPolarisData ExperimentHub package provides two large multiplex immunofluorescence datasets collected by Akoya Biosciences Vectra 3 and Vectra Polaris platforms. Image preprocessing (cell segmentation and phenotyping) was performed using  Inform software. Data cover are formatted into objects of class SpatialExperiment.
vignette: |
  %\VignetteIndexEntry{VectraPolarisData}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---


```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

```{r setup}
library(VectraPolarisData)
```


# multiplex immunofluorescence imaging data

## Available datasets

List the two datasets here and mention that you have detailed data dictionaries below. Give a very brief overview of multiplex imaging data.



# Loading the data

To retrieve a dataset, we can use a dataset's corresponding named function `<id>()`, where `<id>` should correspond to one a valid dataset identifier (see `?TENxVisiumData`). E.g.:

```{r}
library(TENxVisiumData)
spe <- HumanHeart()
```

Alternatively, data can loaded directly from Bioconductor's `r Biocpkg("ExerimentHub")` as follows. First, we initialize a hub instance and store the complete list of records in a variable `eh`. Using `query()`, we then identify any records made available by the `TENxVisiumData` package, as well as their accession IDs (EH1234). Finally, we can load the data into R via `eh[[id]]`, where `id` corresponds to the data entry's identifier we'd like to load. E.g.:

```{r}
library(ExperimentHub)
eh <- ExperimentHub()        # initialize hub instance
q <- query(eh, "TENxVisium") # retrieve 'TENxVisiumData' records
id <- q$ah_id[1]             # specify dataset ID to load
spe <- eh[[id]]              # load specified dataset
```


# Data representation

Talk about spatial experiment

# Data dictionaries

Detailed tables representing what is provided in each dataset are listed here


# Transforming VectraPolarosData to other formats

Show how to transform from a spatial experiment object to a dataframe, if that is your prefered analysis method.

# Citation info

The objects provided in this package are rich data sources we encourage others to use in their own analyses. If you do include them in your peer-reviewed work, we ask that you cite our package and the original studies (either the lung cancer or ovarian cancer study, depending on which data source you use). 

To cite the latest version of `VectraPolarisData` with `bibtex`, use

```{}
# need to add a citation file, see link below

#https://bioconductor.org/developers/package-guidelines/#citation
```

To cite the `HumanLungCancerV3()` data, use:

```{}
@article{johnson2021cancer,
  title={Cancer cell-specific MHCII expression as a determinant of the immune infiltrate organization and function in the non-small cell lung cancer tumor microenvironment},
  author={Johnson, Amber M and Boland, Jennifer M and Wrobel, Julia and Klezcko, Emily K and Weisner-Evans, Mary and Heasley, Lynn and Clambey, Eric T and Nemenoff, Raphael A and Schenk, Erin L},
  journal={bioRxiv},
  year={2021},
  publisher={Cold Spring Harbor Laboratory}
}
```

To cite the `HumanOvarianCancerVP()` data, use:

```{}
@article{steinhart2021spatial,
  title={The spatial context of tumor-infiltrating immune cells associates with improved ovarian cancer survival},
  author={Steinhart, Benjamin and Jordan, Kimberly R and Bapat, Jaidev and Post, Miriam D and Brubaker, Lindsay W and Bitler, Benjamin G and Wrobel, Julia},
  journal={Molecular Cancer Research},
  volume={19},
  number={12},
  pages={1973--1979},
  year={2021},
  publisher={AACR}
}
```



# other analyses using this data

